---
title: "R Notebook"
output: html_notebook
---

# MS&E 125 Final Project
## Grace, Chanse, Carlene, Katelin, Vlad

```{r}
library(tidyverse)
library(readr)
library(dplyr)
library("data.table") 
```

Both cleaned -K
```{r}
#Load in credit recovery data here
credit_recovery <- read_csv("Credit Recovery.csv")
#drop all other states, NA and -9 values
credit_recovery <- credit_recovery %>% filter(LEA_STATE == "CA") %>%  na_if(-9) %>% na_if(-3) %>% na_if(-11) %>%  na_if(-5) %>% na_if(-6) %>% na_if(-8) %>% drop_na()
#rename columns
credit_recovery = rename(credit_recovery, State_Abbreviation = LEA_STATE, State = LEA_STATE_NAME, District = LEA_NAME, School = SCH_NAME, Credit_Recovery_Indicator = SCH_CREDITRECOVERY_IND, Total_Recovery = SCH_CREDITRECOVERYENR)
#drop irrelevant columns
credit_recovery = subset(credit_recovery, select = -c(LEAID, SCHID, COMBOKEY, JJ, State))
#head/summary
head(credit_recovery,15)
summary(credit_recovery)

```
```{r}
#Load in dual enrollment data here
d_enrollment <- read_csv("Dual Enrollment.csv")
#drop all other states, -9 or NA values
#-9 | -3 | -11 | -5 | -6 | -8
d_enrollment <- d_enrollment %>% filter(LEA_STATE == "CA") %>%  
na_if(-9) %>% na_if(-3) %>% na_if(-11) %>%  na_if(-5) %>% na_if(-6) %>% na_if(-8) %>% drop_na()
#rename columns
d_enrollment = rename(d_enrollment, State_Abbreviation = LEA_STATE, State = LEA_STATE_NAME, District = LEA_NAME, School = SCH_NAME, Dual_Indicator = SCH_DUAL_IND, DEMale_Total = TOT_DUAL_M, DEFemale_Total = TOT_DUAL_F)
#drop irrelevant columns
d_enrollment = subset(d_enrollment, select = c(State_Abbreviation, District, School, Dual_Indicator, DEMale_Total, DEFemale_Total, SCH_DUALENR_LEP_M, SCH_DUALENR_LEP_F, SCH_DUALENR_IDEA_M, SCH_DUALENR_IDEA_F))
d_enrollment
summary(d_enrollment)
```

Cleaned -K
```{r}
#load in gifted and talented participation here

#read in only relevant columns
gift_talent = fread("Gifted and Talented.csv", select = c("LEA_STATE", "LEA_NAME", "SCH_NAME", "SCH_GT_IND", "TOT_GTENR_M", "TOT_GTENR_F"))

#rename columns
gift_talent = rename(gift_talent, State_Abbreviation = LEA_STATE, District = LEA_NAME, School = SCH_NAME, G_T_Indicator = SCH_GT_IND, GTMale_Total = TOT_GTENR_M, GTFemale_Total = TOT_GTENR_F)

#filter out all states that aren't CA
gift_talent = gift_talent[gift_talent$State_Abbreviation == "CA", ]

#remove -9 or NA

gift_talent[gift_talent < 0] <- NA

gift_talent = drop_na(gift_talent)
gift_talent = gift_talent[gift_talent$GTMale_Total != -9,]
gift_talent = gift_talent[gift_talent$GTFemale_Total != -9,]

summary(gift_talent)
gift_talent
```

Cleaned -K
```{r}
# load in SAT participation here

#read in SAT/ACT data, only read in necessary columns
SAT_ACT = fread("SAT and ACT.csv", select=c("LEA_STATE","LEA_STATE_NAME","LEA_NAME","SCH_NAME","TOT_SATACT_M","TOT_SATACT_F"))

#filter out all states that aren't CA
SAT_ACT = SAT_ACT[SAT_ACT$LEA_STATE == "CA", ]

#rename columns
SAT_ACT = rename(SAT_ACT, State_Abbreviation = LEA_STATE, State = LEA_STATE_NAME, District = LEA_NAME, School = SCH_NAME, SA_Male = TOT_SATACT_M, SA_Female = TOT_SATACT_F)

SAT_ACT[SAT_ACT < 0] <- NA
SAT_ACT = drop_na(SAT_ACT)
summary(SAT_ACT)
SAT_ACT
```

Cleaned -K
#Comment: Uh there is very few data points here (97). There may be,,, none,,, when we do the joins..
```{r}
#Load in IB Data here
IB <- read_csv("International Baccalaureate.csv")
#drop all other states, -9 or NA values
IB <- IB %>% filter(LEA_STATE == "CA") %>%  na_if(-9) %>% na_if(-3) %>% na_if(-11) %>%  na_if(-5) %>% na_if(-6) %>% na_if(-8) %>% drop_na()
#rename columns
IB = rename(IB, State_Abbreviation = LEA_STATE, State = LEA_STATE_NAME, District = LEA_NAME, School = SCH_NAME, IB_Indicator = SCH_IBENR_IND, Male_Total = TOT_IBENR_M, Female_Total = TOT_IBENR_F)
#drop irrelevant columns
IB = subset(IB, select = c(State_Abbreviation, State, District, School, IB_Indicator, Male_Total, Female_Total))
#summary/head
summary(IB)
head(IB,15)
```
Cleaned -K
```{r}
#load in school characteristics

#read in school characteristics, only read in necessary columns
characteristics = fread("School Characteristics.csv", select = c("LEA_STATE","LEA_STATE_NAME","LEA_NAME","SCH_NAME","SCH_GRADE_G09", "SCH_GRADE_G10", "SCH_GRADE_G11", "SCH_GRADE_G12"))

#filter out all states that aren't CA
characteristics = characteristics[characteristics$LEA_STATE == "CA", ]

#rename columns
characteristics= rename(characteristics, State_Abbreviation = LEA_STATE, State = LEA_STATE_NAME, District = LEA_NAME, School = SCH_NAME, Grades_Stu_Enrolled_9 = SCH_GRADE_G09, Grades_Stu_Enrolled_10 = SCH_GRADE_G10, Grades_Stu_Enrolled_11 = SCH_GRADE_G11, Grades_Stu_Enrolled_12 = SCH_GRADE_G12)

#drop anything that has NA
characteristics = drop_na(characteristics)
summary(characteristics)

characteristics 
```

Cleaned -K
```{r}
#Load in expenditures data (including federal, state, local funding)
expenditures = fread("School Expenditures.csv", select = c("LEA_STATE","LEA_STATE_NAME","LEA_NAME","SCH_NAME","SCH_SAL_TEACH_WFED", "SCH_SAL_ADM_WFED", "SCH_SAL_TOTPERS_WFED", "SCH_NPE_WFED"))

#filter out all states that aren't CA
expenditures = expenditures[expenditures$LEA_STATE == "CA", ]

#rename columns 
expenditures = rename(expenditures, State_Abbreviation = LEA_STATE, State = LEA_STATE_NAME, District = LEA_NAME, School = SCH_NAME, Teacher_Expenditures = SCH_SAL_TEACH_WFED, Admin_Expenditures = SCH_SAL_ADM_WFED, All_Staff_Expenditures = SCH_SAL_TOTPERS_WFED, Activity_Expenditures = SCH_NPE_WFED)

#drop anything that has NA or -9
expenditures = na_if(expenditures, -9)
expenditures = na_if(expenditures, 0)

expenditures = drop_na(expenditures)
summary(expenditures)

head(expenditures,15)
# expenditure_model <- lm(Teacher_Expenditures ~ Activity_Expenditures, 
#                   data = expenditures,
#                   family = binomial)
#summary(expenditure_model)
```

```{r}
# Chanse's exploration
# college_going_1 <- college_going %>% 
#   mutate_at(c(13:20), as.numeric) %>%
#    filter(!is.na(College.Going.Rate...Total..12.Months.)) %>%
#    mutate(CharterSchool = (CharterSchool == "Yes"))
# 
# charter_schools <- college_going_1 %>% filter(CharterSchool == TRUE)
# noncharter_schools <- college_going_1 %>% filter(CharterSchool == FALSE)
# mean(charter_schools$College.Going.Rate...Total..12.Months.)
# mean(noncharter_schools$College.Going.Rate...Total..12.Months.)
```

```{r}
# Grace clean up college_going 

college_going = read.delim("collegegoing.txt", na.strings=c("","*", "NA"))

college_going <- college_going %>% select(AggregateLevel, CountyCode, DistrictCode, SchoolCode, DistrictName, SchoolName, CharterSchool, ReportingCategory, CompleterType, High.School.Completers, Enrolled.In.College...Total..12.Months., College.Going.Rate...Total..12.Months., Enrolled.In.State..12.Months.,Enrolled.Out.of.State..12.Months.) 

college_going <- college_going %>% rename(School = SchoolName, District = DistrictName, Total_Enrolled_College = Enrolled.In.College...Total..12.Months., Total_College_Going_Rate = College.Going.Rate...Total..12.Months., Total_Enrolled_In_State = Enrolled.In.State..12.Months., Total_Enrolled_Out_State = Enrolled.Out.of.State..12.Months.)

college_going <- na.omit(college_going)
college_going <- college_going %>% distinct()
college_going <- filter(college_going, CompleterType == 'TA')

#KATELIN CLEANING

#filter only by school name
schools_only = college_going[college_going$AggregateLevel == "S", ]
#filter only by total student count (no breakdown by race/ethnicity)
schools_only = schools_only[schools_only$ReportingCategory == "TA", ]
#filter only by all schools, regardless of charter status
schools_only = schools_only[schools_only$CharterSchool == "All", ]
#delete all columns no longer necessary (all listed above)
schools_only = subset(schools_only, select = -AggregateLevel)
schools_only = subset(schools_only, select = -ReportingCategory)
schools_only = subset(schools_only, select = -CharterSchool)
schools_only = subset(schools_only, select = -CompleterType)
#remove all duplicate lines
schools_only = schools_only[!duplicated(schools_only), ]

schools_only = subset(schools_only, Total_Enrolled_College != 0)
summary(schools_only)
head(schools_only,15)
```

```{r}

```

#katelin changes 5/31 2:38 PM 
- Added back aggregate level to sort by schools only (eliminating district/state/county designation)
- Added back reporting category to display only totals across races (no breakdown)
- Remove charter school designation to display all (regardless of charter status) - eliminates duplicates

```{r}
# Grace load in graduation rate
grad_rate = read.delim("graduationrate.txt", na.strings=c("","*", "NA"))
grad_rate <- grad_rate %>% 
select(AggregateLevel, CountyCode, DistrictCode, SchoolCode, DistrictName, SchoolName, CharterSchool, DASS, ReportingCategory, CohortStudents, Regular.HS.Diploma.Graduates..Count.,Regular.HS.Diploma.Graduates..Rate.) %>% rename(School=SchoolName, District=DistrictName,Num_Students = CohortStudents, HS_Diploma_Count = Regular.HS.Diploma.Graduates..Count., HS_Diploma_Rate = Regular.HS.Diploma.Graduates..Rate.)
grad_rate <- na.omit(grad_rate)

#katelin edits

#filter only by school name
grad_rate_schools = grad_rate[grad_rate$AggregateLevel == "S", ]

#filter by charter school, DASS, and reporting category
grad_rate_schools = grad_rate_schools[grad_rate_schools$ReportingCategory == "TA", ]
grad_rate_schools = grad_rate_schools[grad_rate_schools$DASS == "All", ]
grad_rate_schools = grad_rate_schools[grad_rate_schools$CharterSchool == "All", ]

#remove unnecessary columns
grad_rate_schools = subset(grad_rate_schools, select = -c(ReportingCategory, AggregateLevel, DASS, CharterSchool))
grad_rate_schools = grad_rate_schools[!duplicated(grad_rate_schools), ]

grad_rate_schools = subset(grad_rate_schools, HS_Diploma_Rate != 0)
grad_rate_schools


#grad_test <- grad_rate_schools %>% -filter(HS_Diploma_Rate == 0.0)
#grad_test
```

```{r}
# use inner join to merge all the cleaned up data files 
joined_data <- inner_join(schools_only, grad_rate_schools, by=c("DistrictCode", "SchoolCode", "CountyCode", "District", "School"))
joined_data
```

```{r}
diploma_college_model <- lm(formula = Total_College_Going_Rate ~ 1 + HS_Diploma_Rate, data = joined_data)
summary(diploma_college_model)

ggplot(diploma_college_model, aes(x=HS_Diploma_Rate, y=Total_College_Going_Rate)) +  
    geom_point() + 
    geom_smooth(method="lm", formula = y ~ x) + ggtitle("Correlation between HS Diploma Rate and College Going Rate") + xlab("High School Diploma Rate") +ylab("Total College Going Rate")

ggsave("diploma_college_model.png")
```

```{r}
# TODO: Run regression analysis of expenditures and high school grad / college going rates
expenditure_regression <- inner_join(joined_data, expenditures, by=c("School", "District"))
expenditure_regression = select(expenditure_regression, -c(State,State_Abbreviation))
summary(expenditure_regression)
expenditure_regression

ggplot(expenditure_regression, aes(x=All_Staff_Expenditures)) + geom_histogram(binwidth = 100000) 

new_model <- lm(formula = Total_College_Going_Rate ~ 1 +  All_Staff_Expenditures, data = expenditure_regression)
summary(new_model)

ggplot(expenditure_regression, aes(x=All_Staff_Expenditures, y=Total_College_Going_Rate)) +  
    geom_point() + 
    geom_smooth(method="lm", formula = y ~ x) +
    geom_smooth(method="lm", formula = y ~ I(log(x)))


ex_model <- lm(formula = HS_Diploma_Rate ~ 1 +  All_Staff_Expenditures, data = expenditure_regression)
summary(ex_model)

ggplot(expenditure_regression, aes(x=All_Staff_Expenditures, y=HS_Diploma_Rate)) +  
    geom_point() + 
    geom_smooth(method="lm", formula = y ~ x) +
    geom_smooth(method="lm", formula = y ~ I(log(x)))

ST_model <- lm(formula = All_Staff_Expenditures ~ 1 + Num_Students , data = expenditure_regression)
summary(ST_model)

ggplot(expenditure_regression, aes(x=Num_Students, y=All_Staff_Expenditures)) +  
    geom_point() + 
    geom_smooth(method="lm", formula = y ~ x) +
    geom_smooth(method="lm", formula = y ~ I(log(x)))


pred_253kids <- predict(ST_model, newdata = tibble(Num_Students = mean(expenditure_regression$Num_Students)))
# avg number of students is 253
pred_253kids  # 253 kids is about $5338394 million in staff expenditures
# Not sure if need to split up by test/train set, one method could be sampling a few data points from each county code?

```

```{r}
# TODO: Run regression analysis of SAT_ACT participation  and high school grad/college going rates

SAT_ACT_regression <- inner_join(joined_data, SAT_ACT, by=c("School", "District"))
SAT_ACT_regression = select(SAT_ACT_regression, -c(State,State_Abbreviation))

SAT_model <-  lm(formula = Total_College_Going_Rate ~ 1 + SA_Male + SA_Female, data=SAT_ACT_regression)

ggplot(SAT_ACT_regression, aes(x=SA_Male, y=Total_College_Going_Rate)) +  
    geom_point() + 
    geom_smooth(method="lm", formula = y ~ x) 

summary(SAT_model)

SAT_model_d <-  lm(formula = HS_Diploma_Rate ~ 1 + SA_Male + SA_Female, data=SAT_ACT_regression)

summary(SAT_model_d)


```

```{r}
# TODO: Run regression analysis of SAT_ACT participation, GT program and high school grad/college going rates

SAT_ACT_regression <- inner_join(joined_data, SAT_ACT, by=c("School", "District"))
SAT_ACT_regression = select(SAT_ACT_regression, -c(State,State_Abbreviation))

#joining gifted/talented
gift_talent_r <- inner_join(SAT_ACT_regression, gift_talent, by=c("District","School"))
gift_talent_r <- select(gift_talent_r, -State_Abbreviation)

SAT_gift_model <- lm(formula = Total_College_Going_Rate ~ 1 + SA_Male + SA_Female + GTMale_Total + GTFemale_Total, data=gift_talent_r)

summary(SAT_gift_model)

SAT_gift_model_d <- lm(formula = HS_Diploma_Rate ~ 1 + SA_Male + SA_Female + GTMale_Total + GTFemale_Total, data=gift_talent_r)

summary(SAT_gift_model_d)

```

```{r}
# TODO: Run regression analysis of characteristics and high school grad / college going rates
char_regression <- inner_join(joined_data, characteristics, by=c("School", "District"))
char_regression = select(char_regression, -c(State,State_Abbreviation))
summary(char_regression)
char_regression

# Not sure if need to split up by test/train set, one method could be sampling a few data points from each county code?
# interesting thing is that we need to treat this as binary variable?? because there's "yes" vs "no" data so maybe this is not a good regression lol

```

```{r}
# biggest big boi regression here 
# join credit recovery, dual enrollment, gifted/talented, SAT/ACT participation, expenditures

#joining dual enrollment
d_data <- inner_join(joined_data,d_enrollment, by=c("District","School"))
d_data = select(d_data, -State_Abbreviation)

#joining SAT_ACT participation
SAT <- inner_join(d_data, SAT_ACT,  by=c("District","School"))
SAT = select(SAT, -c(State,State_Abbreviation))

#joining gifted/talented
GT <- inner_join(SAT, gift_talent, by=c("District","School"))
GT = select(GT, -State_Abbreviation)

#joining credit recovery
CR <- inner_join(GT, credit_recovery,  by=c("District","School"))
CR = select(CR, -State_Abbreviation)

#joining expenditures
EXP <- inner_join(CR, expenditures, by=c("District","School"))
EXP = select(EXP, -c(State_Abbreviation,State))

total_data <- EXP
head(total_data,10)

complete_model <- lm(formula = Total_College_Going_Rate ~ 1 + HS_Diploma_Rate + DEMale_Total + DEFemale_Total + SA_Male + SA_Female + GTMale_Total + GTFemale_Total + Total_Recovery + Teacher_Expenditures + All_Staff_Expenditures + Admin_Expenditures + Activity_Expenditures, data=total_data)

summary(complete_model)

```

