---
title: "R Notebook"
output: html_notebook
---

# MS&E 125 Final Project
## Grace, Chanse, Carlene, Katelin, Vlad

```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(stringi)
library("data.table") 
```


```{r}
#load in school characteristics
#read in school characteristics, only read in necessary columns
characteristics = fread("https://raw.githubusercontent.com/graceclee/125project/main/School%20Characteristics.csv", select = c("LEA_STATE","LEA_STATE_NAME","LEA_NAME","SCH_NAME","SCH_GRADE_G09", "SCH_GRADE_G10", "SCH_GRADE_G11", "SCH_GRADE_G12"))
#filter out all states that aren't CA
characteristics = characteristics[characteristics$LEA_STATE == "CA", ]
#rename columns
characteristics= rename(characteristics, State_Abbreviation = LEA_STATE, State = LEA_STATE_NAME, District = LEA_NAME, School = SCH_NAME, Grades_Stu_Enrolled_9 = SCH_GRADE_G09, Grades_Stu_Enrolled_10 = SCH_GRADE_G10, Grades_Stu_Enrolled_11 = SCH_GRADE_G11, Grades_Stu_Enrolled_12 = SCH_GRADE_G12)
#drop anything that has NA
characteristics = drop_na(characteristics)
summary(characteristics)
characteristics 
```

```{r}
#Load in expenditures data (including federal, state, local funding)
expenditures = fread("https://github.com/graceclee/125project/blob/main/School%20Expenditures.csv?raw=true", select = c("LEA_STATE","LEA_STATE_NAME","LEA_NAME","SCH_NAME","SCH_SAL_TEACH_WFED", "SCH_SAL_ADM_WFED", "SCH_SAL_TOTPERS_WFED", "SCH_NPE_WFED"))
#filter out all states that aren't CA
expenditures = expenditures[expenditures$LEA_STATE == "CA", ]
#rename columns 
expenditures = rename(expenditures, State_Abbreviation = LEA_STATE, State = LEA_STATE_NAME, District = LEA_NAME, School = SCH_NAME, Teacher_Expenditures = SCH_SAL_TEACH_WFED, Admin_Expenditures = SCH_SAL_ADM_WFED, All_Staff_Expenditures = SCH_SAL_TOTPERS_WFED, Activity_Expenditures = SCH_NPE_WFED)
#drop anything that has NA or -9
expenditures = na_if(expenditures, -9)
expenditures = na_if(expenditures, 0)
expenditures = drop_na(expenditures)
summary(expenditures)
head(expenditures,15)

# expenditure_model <- lm(Teacher_Expenditures ~ Activity_Expenditures, 
#                   data = expenditures,
#                   family = binomial)
#summary(expenditure_model)

```

Cleaned -K
```{r}
# load in SAT participation here
#read in SAT/ACT data, only read in necessary columns
SAT_PATH <- "https://raw.githubusercontent.com/graceclee/125project/main/SAT%20and%20ACT.csv"
SATdata <- read.csv(SAT_PATH)
CA_DATA <- SATdata %>% 
  filter(LEA_STATE == "CA") 

allnines = -9 * 20

SAT_ACT <- CA_DATA %>%
  filter(rowSums(CA_DATA[,9:28])!= allnines) %>% 
  select(LEA_NAME,  SCH_NAME,TOT_SATACT_M,  TOT_SATACT_F)

SAT_ACT = rename(SAT_ACT, District = LEA_NAME, School = SCH_NAME,  Female_Total = TOT_SATACT_F, Male_Total = TOT_SATACT_M)


SAT_ACT[SAT_ACT < 0] <- NA


SAT_ACT
```




```{r}
# Chanse's exploration
# college_going_1 <- college_going %>% 
#   mutate_at(c(13:20), as.numeric) %>%
#    filter(!is.na(College.Going.Rate...Total..12.Months.)) %>%
#    mutate(CharterSchool = (CharterSchool == "Yes"))
# 
# charter_schools <- college_going_1 %>% filter(CharterSchool == TRUE)
# noncharter_schools <- college_going_1 %>% filter(CharterSchool == FALSE)
# mean(charter_schools$College.Going.Rate...Total..12.Months.)
# mean(noncharter_schools$College.Going.Rate...Total..12.Months.)
```

```{r}
# Grace clean up college_going 
college_going = read.delim("https://github.com/graceclee/125project/blob/main/collegegoing.txt?raw=true", na.strings=c("","*", "NA"))
college_going <- college_going %>% select(AggregateLevel, CountyCode, DistrictCode, SchoolCode, DistrictName, SchoolName, CharterSchool, ReportingCategory, CompleterType, High.School.Completers, Enrolled.In.College...Total..12.Months., College.Going.Rate...Total..12.Months., Enrolled.In.State..12.Months.,Enrolled.Out.of.State..12.Months.) 
college_going <- college_going %>% rename(School = SchoolName, District = DistrictName, Total_Enrolled_College = Enrolled.In.College...Total..12.Months., Total_College_Going_Rate = College.Going.Rate...Total..12.Months., Total_Enrolled_In_State = Enrolled.In.State..12.Months., Total_Enrolled_Out_State = Enrolled.Out.of.State..12.Months.)
college_going <- na.omit(college_going)
college_going <- college_going %>% distinct()
college_going <- filter(college_going, CompleterType == 'TA')
#KATELIN CLEANING
#filter only by school name
schools_only = college_going[college_going$AggregateLevel == "S", ]
#filter only by total student count (no breakdown by race/ethnicity)
schools_only = schools_only[schools_only$ReportingCategory == "TA", ]
#filter only by all schools, regardless of charter status
schools_only = schools_only[schools_only$CharterSchool == "All", ]
#delete all columns no longer necessary (all listed above)
schools_only = subset(schools_only, select = -AggregateLevel)
schools_only = subset(schools_only, select = -ReportingCategory)
schools_only = subset(schools_only, select = -CharterSchool)
schools_only = subset(schools_only, select = -CompleterType)
#remove all duplicate lines
schools_only = schools_only[!duplicated(schools_only), ]
schools_only = subset(schools_only, Total_Enrolled_College != 0)
summary(schools_only)
schools_only
```


#katelin changes 5/31 2:38 PM 
- Added back aggregate level to sort by schools only (eliminating district/state/county designation)
- Added back reporting category to display only totals across races (no breakdown)
- Remove charter school designation to display all (regardless of charter status) - eliminates duplicates

```{r}
# Grace load in graduation rate
grad_rate = read.delim("https://github.com/graceclee/125project/blob/main/graduationrate.txt?raw=true", na.strings=c("","*", "NA"))
grad_rate
grad_rate <- grad_rate %>% 
select(AggregateLevel, CountyCode, DistrictCode, SchoolCode, DistrictName, SchoolName, CharterSchool, DASS, ReportingCategory, CohortStudents, Regular.HS.Diploma.Graduates..Count.,Regular.HS.Diploma.Graduates..Rate.) %>% rename(School=SchoolName, District=DistrictName,Num_Students = CohortStudents, HS_Diploma_Count = Regular.HS.Diploma.Graduates..Count., HS_Diploma_Rate = Regular.HS.Diploma.Graduates..Rate.)
grad_rate <- na.omit(grad_rate)
#katelin edits
#filter only by school name
grad_rate_schools = grad_rate[grad_rate$AggregateLevel == "S", ]
#filter by charter school, DASS, and reporting category
grad_rate_schools = grad_rate_schools[grad_rate_schools$ReportingCategory == "TA", ]
grad_rate_schools = grad_rate_schools[grad_rate_schools$DASS == "All", ]
grad_rate_schools = grad_rate_schools[grad_rate_schools$CharterSchool == "All", ]
#remove unnecessary columns
grad_rate_schools = subset(grad_rate_schools, select = -c(ReportingCategory, AggregateLevel, DASS, CharterSchool))
grad_rate_schools = grad_rate_schools[!duplicated(grad_rate_schools), ]
grad_rate_schools = subset(grad_rate_schools, HS_Diploma_Rate != 0)
grad_rate_schools
#grad_test <- grad_rate_schools %>% -filter(HS_Diploma_Rate == 0.0)
#grad_test
```

```{r}
# use inner join to merge all the cleaned up data files 
joined_data <- inner_join(schools_only, grad_rate_schools, by=c("DistrictCode", "SchoolCode", "CountyCode", "District", "School"))
joined_data
# small_schools_joined <- joined_data %>% 
#   filter(Num_Students <= 75)
joined_data_withschoolsize <- joined_data %>%
  mutate(size = cut_number(Num_Students, n = 3, labels = c("small", "medium","large"))) 
joined_data_withschoolsize
```

```{r}
diploma_college_model <- lm(formula = Total_College_Going_Rate ~ 1 + HS_Diploma_Rate, data = joined_data_withschoolsize)
summary(diploma_college_model)
ggplot(joined_data_withschoolsize, aes(x=HS_Diploma_Rate, y=Total_College_Going_Rate, , color = size)) +
    geom_point() + 
    geom_smooth(method="lm", formula = y ~ x) + ggtitle("Correlation between HS Diploma Rate and College Going Rate") + xlab("High School Diploma Rate") +ylab("Total College Going Rate")
```

```{r}
#joining the SAT and school data 
SAT_and_grad_rate <- inner_join(SAT_ACT, grad_rate_schools, by=c("District", "School"))  %>% 
  select(District, School, Male_Total, Female_Total, Num_Students) %>%
  #key assumption most likely num students not updated properly, transfers etc. 
  mutate(test_part_rate = ifelse(Male_Total + Female_Total > Num_Students, 100, (Male_Total + Female_Total)/Num_Students * 100) ) %>% 
  mutate(size = cut_number(Num_Students, n = 3, labels = c("small", "medium","large"))) 
  

cleaned_SAT_and_grad_rate <- inner_join(SAT_and_grad_rate, schools_only, by=c("District", "School")) %>% 
   select(District, School, test_part_rate, Total_College_Going_Rate, size)


cleaned_SAT_and_grad_rate

```
```{r}
college_sat_model <- lm(formula = Total_College_Going_Rate ~ 1 + test_part_rate, data = cleaned_SAT_and_grad_rate)
summary(college_sat_model)
ggplot(cleaned_SAT_and_grad_rate, aes(x=test_part_rate, y=Total_College_Going_Rate, color = size)) +
    geom_point(alpha = .6) + 
    geom_smooth(method="lm", formula = y ~ x) + ggtitle("college participation based on school SAT part rate") + xlab("SAT part rate") +ylab("Total College Going Rate")
```

```{r}
# expenditure data with college goiong rate 
cleaned_expenditures <- expenditures %>% 
  mutate(total_expenditure = All_Staff_Expenditures + Activity_Expenditures) 

joined_wealth <- inner_join(cleaned_expenditures, grad_rate_schools, by=c("District", "School"))  %>% 
  select(District, School, total_expenditure, Num_Students) %>% 
  mutate(school_income_per_student = total_expenditure / (Num_Students * 4)) %>% 
  mutate(school_income_per_student = cut_number(total_expenditure, n = 3, labels = c("lower", "middle","upper")))

joined_wealth_final <- inner_join(joined_wealth, joined_data, by=c("District", "School"))
summary(joined_wealth_final)

ggplot(joined_wealth_final, aes(x=HS_Diploma_Rate, y=Total_College_Going_Rate, , color = school_income_per_student)) +
    geom_point() + 
    geom_smooth(method="lm", formula = y ~ x) + ggtitle("HS Diploma Rate and College Going Rate by school wealth per student") + xlab("High School Diploma Rate") +ylab("Total College Going Rate")
```

```{r}
#sat part rate on college going rate, data broken down by wealth 

cleaned_SAT_and_grad_rate_wealth <- inner_join(cleaned_SAT_and_grad_rate, joined_wealth_final, by=c("District", "School"))
summary(cleaned_SAT_and_grad_rate_wealth)

ggplot(cleaned_SAT_and_grad_rate_wealth, aes(x=test_part_rate, y=Total_College_Going_Rate.x, color = school_income_per_student)) +
    geom_point(alpha = .6) + 
    geom_smooth(method="lm", formula = y ~ x) + ggtitle("college participation based on school SAT part rate with school expenditure") + xlab("SAT part rate") +ylab("Total College Going Rate")
```
    
